title

### 2023-11-01 outputs/paper/paper.qmd
LR model section improved. Using original data instead of downsampled since performance is way better

---
title: "title"
subtitle: "subtitle"
author: 
  - Michaela Drouillard
thanks: "Code and data are available at: https://github.com/michaeladrouillard/jack."
date: today
abstract: "Using data from the Spotify API, I explored Jack Antonoff's production style and assessed whether Spotify's audio features can capture it. In an interview with Glen McDonald, an engineer at Spotify, I learned that the danceability and valence features were created by having interns tag songs as positive or gloomy, or danceable or un-danceable. Using a logistic regression model, I found that danceability was the most important feature in predicting Antonoff's involvement in a song. In the discussion section, I do a close-listen of Antonoff's most danceable songs, and argue that the danceability variable is actually capturing something more like a visceral quality to Antonoff's music."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

I made a binary response variable, called `jack`, that has a value of 1 if the song was written by Jack, and a 0 otherwise. There are 10 predictor variables, all of which are continuous: **`danceability`**, **`energy`**, **`loudness`**, **`mode`**, **`speechiness`**, **`acousticness`**, **`instrumentalness`**, **`liveness`**, **`valence`**, and **`tempo`**.

@@

where $\text{Pr}(y_i=1)$ is the probability of a song being written by Jack, $\beta_0$ is the intercept, and $\beta_1, \ldots, \beta_{10}$ are the coefficients corresponding to the predictor variables.

@fig-means gives us an idea of the mean value of each of these features for Jack and non-Jack song. A logistic regression will give us a better idea of how to weigh each of these features in order to predict the outcome, which gives us information about the "research" "question": are these characteristics related to Jack's music? (And then: Which ones? And then, the most elusive question: How?)

@@

data_reduced |>
  group_by(jack) |>
  summarise(across(-c(tempo, loudness), mean)) |>
  pivot_longer(cols = -jack) |>
  ggplot(aes(value, name, fill = factor(jack))) +
  geom_col(alpha = 0.8, position = "dodge") +
  theme_minimal() +
  labs(fill = "Jack", x = "Mean Value", y = "Feature")
```

This model assumes that the relationship between the independent variables and the dependent variable is linear. There could be non-linear relationships between the variables that this model isn't capturing. Second, it assumes that the relationship between the independent variables and the dependent variable is the same for all observations, which might not be true in practice -- there could be interactions between the variables, or in certain subgroups of data where the relationship is different. Albums where Jack may have had more of a back seat producing role are one example. (The fact that a song's development, from inception to recording, is a whole creative process, each having its own origin story with varying degrees of Jack-ness, is another.)

@@

I used **`rank_results()`** from the **`yardstick`** package to rank the models based on accuracy, log loss, sensitivity, and specificity, and went with the downsampled model.

@@

The model's accuracy score is 0.743, and the ROC AUC score is 0.841. The accuracy score could be misleading, since the model had imbalanced classes, and downsampling may have affected the model's ability to correctly identify the minority class. The ROC AUC score is more reliable, since it's less sensitive to class imbalance.

|                        | Actually Positive (1) | Actually Negative (0) |
|------------------------|-----------------------|-----------------------|
| Predicted Positive (1) | 581                   | 25                    |
| Predicted Negative (0) | 194                   | 51                    |

@@

```{r}
#| label: tbl-importance
#| tbl-cap: Importance Scores of Each Variable in Determining Jack-ness
#| echo: false
#| warning: false

jack_vip <- read_csv(here::here("inputs/data/jack_vip.csv"))

jack_vip |>
  knitr::kable(col.names = c("Variable",
                             "Importance",
                             "Sign"))
```

```{r}
#| label: fig-importance
#| fig-cap: Importance Scores of Each Variable in Determining Jack-ness
#| echo: false
#| warning: false

jack_vip |>
  group_by(Sign) |>
  slice_max(Importance, n = 15) |>
  ungroup() |>
  ggplot(aes(Importance, fct_reorder(Variable, Importance), fill = Sign)) +
  geom_col() +
  facet_wrap(vars(Sign), scales = "free_y") +
  labs(y = NULL) +
  theme(legend.position = "none")
```

Danceability, with a positive importance score of 4.32, had the highest level of importance in determining whether a song was produced or co-produced by Jack Antonoff. Instrumentalness comes in second at 2.59, with a negative score (meaning that negative instrumentalness is correlated), and valence comes in third with a negative score at 2.02.

@@

df |>
  filter(jack == 1) |>
  arrange(desc(danceability)) |>
  distinct(track_name, .keep_all = TRUE) |>
  select(artist_name, danceability, track_name) |>
  head(20) |>
  knitr::kable(col.names = c("Artist Name",
                             "Danceability",
                             "Track"))
```

@@

The categorical feature 'producer', which contains the names of the producer for each track, was one-hot encoded to create binary columns for each producer ('is_antonoff', 'is_elworth', etc.). The producer feature was then converted into a factor variable, and the dataset was split into training and tests sets.

@@

The RF was trainied using the @ranger package. The 'importance' parameter was set to 'impurity', which sets the model to interpret variable importance using Gini impurity.

@@

Gini impurity was chose on the grounds that this study is interested in not only classifying producers, but, more importantly, in understanding how the underlying data informs these decisions. By calculating the Gini impurity not only for the entire model, but specifically for each producer, we can better understand how the variables in our dataset describe the differences between producers. Without song structure, instruments, sound, or other more ephemeral components of style, these differences can be taken as a proxy for style.

@@

>>>>

---
title: "title"
subtitle: "subtitle"
author: 
  - Michaela Drouillard
thanks: "Code and data are available at: https://github.com/michaeladrouillard/spotify_pop."
date: today
abstract: "Using data from the Spotify API, I explored Jack Antonoff's production style and assessed whether Spotify's audio features can capture it. In an interview with Glen McDonald, an engineer at Spotify, I learned that the danceability and valence features were created by having interns tag songs as positive or gloomy, or danceable or un-danceable. Using a logistic regression model, I found that danceability was the most important feature in predicting Antonoff's involvement in a song. In the discussion section, I do a close-listen of Antonoff's most danceable songs, and argue that the danceability variable is actually capturing something more like a visceral quality to Antonoff's music."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

The binary response variable, called `is_antonoff`, has a value of 1 if the song was produced by Antonoff, and a 0 otherwise. There are 10 predictor variables, all of which are continuous: **`danceability`**, **`energy`**, **`loudness`**, **`mode`**, **`speechiness`**, **`acousticness`**, **`instrumentalness`**, **`liveness`**, **`valence`**, and **`tempo`**.

@@

where $\text{Pr}(y_i=1)$ is the probability of a song being produced by Antonoff, $\beta_0$ is the intercept, and $\beta_1, \ldots, \beta_{10}$ are the coefficients corresponding to the predictor variables.

@fig-means gives us an idea of the mean value of each of these features for Antonofff and non-Antonoff-produced tracks.

@@

lr_data <- read_csv(here::here("inputs/data/lr_data.csv"))

lr_data |>
  group_by(is_antonoff) |>
  summarise(across(-c(tempo, loudness), mean)) |>
  pivot_longer(cols = -is_antonoff) |>
  ggplot(aes(value, name, fill = factor(is_antonoff))) +
  geom_col(alpha = 0.8, position = "dodge") +
  theme_minimal() +
  labs(fill = "Jack", x = "Mean Value", y = "Feature")
```

@@

I used **`rank_results()`** from the **`yardstick`** package to rank the models based on accuracy, log loss, sensitivity, and specificity. After testing both basic and downsampled versions of the model, the basic version yielded much higher accuracy score (0.918 basic versus 0.743 downsampled), although a lower roc auc score (0.746 basic versus 0.821 downsampled).

The version with the original data remains consistently high in accuracy, even with increased regularization. It starts low with in mean log loss and sensitivity, but increases with increased regularization. Compared to the downsampled data, which starts high but drops significantly in accuracy and sensitivity, we chose to use the original data.

![Comparing the Performance for Downsampled and Original Data](images/downsamplevsorig)

@@

The model's accuracy score is 0.918, and the ROC AUC score is 0.746.

|                        | Actually Positive (1) | Actually Negative (0) |
|------------------------|-----------------------|-----------------------|
| Predicted Positive (1) | 848                   | 76                    |
| Predicted Negative (0) | 1                     | 0                     |

@@

```{r}
#| label: tbl-importance
#| tbl-cap: Importance Scores of Each Variable in Determining Antonoff's Tracks
#| echo: false
#| warning: false

antonoff_vip <- read_csv(here::here("inputs/data/antonoff_vip.csv"))

antonoff_vip |>
  knitr::kable(col.names = c("Variable",
                             "Importance",
                             "Sign"))
```

```{r}
#| label: fig-importance
#| fig-cap: Importance Scores of Each Variable in Determining Antonoff's tracks
#| echo: false
#| warning: false

antonoff_vip |>
  group_by(Sign) |>
  slice_max(Importance, n = 15) |>
  ungroup() |>
  ggplot(aes(Importance, fct_reorder(Variable, Importance), fill = Sign)) +
  geom_col() +
  facet_wrap(vars(Sign), scales = "free_y") +
  labs(y = NULL) +
  theme(legend.position = "none")
```

Danceability, with a positive importance score of 4.142, had the highest level of importance in determining whether a song was produced or co-produced by Jack Antonoff. Valence comes in second at 2.559, with negative score. The third, instrumentalness correlates negatively with Antonoff tracks with a a negative score of 2.309.

@@

clean_df <- read_csv(here::here("inputs/data/clean_df.csv"))

clean_df |>
  filter(producer == "antonoff") |>
  arrange(desc(danceability)) |>
  distinct(track_name, .keep_all = TRUE) |>
  select(artist_name, danceability, track_name) |>
  head(20) |>
  knitr::kable(col.names = c("Artist Name",
                             "Danceability",
                             "Track"))
```

@@

The categorical feature `producer`, which contains the names of the producer for each track, was one-hot encoded to create binary columns for each producer (i.e. `is_antonoff`, `is_elworth`, etc.). The producer feature was then converted into a factor variable, and the dataset was split into training and tests sets.

@@

The RF was trained using the @ranger package. The 'importance' parameter was set to 'impurity', which sets the model to interpret variable importance using Gini impurity.

@@

Gini impurity was chosen on the grounds that this study is interested in not only classifying producers, but, more importantly, in understanding how the underlying data informs these decisions. By calculating the Gini impurity not only for the entire model, but specifically for each producer, we can better understand how the variables in our dataset describe the differences between producers. Without song structure, instruments, sound, or other more ephemeral components of style, these differences can be taken as a proxy for style.

@@

The accuracy of the model (0.6111) being significantly higher than the No Information Rate (0.4274) means that it is doing more than just blind guessing based on class prevalence. With a p-value below 0.05, we can confirm the model's statistical significance over a naive approach.

The model's accuracy and Kappa statistics suggest a moderate level of performance. However, there is enough demonstrated statistical significance to suggest that there are underlying patterns in the data that differentiate producers.
